Prehensile Manipulation Planning: Modelling, Algorithms and Implementation

### 2021-09-01 hpp.tex
Take into account reviewers comments.

@@

This \paper therefore describes a software platform called \href{https://humanoid-path-planner.github.io/hpp-doc}{Humanoid Path
  Planner} taylored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modeling of prehensile manipulation based on
  non-linear constraints,
\item an original solver for non-linear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state of the art development tools and methods.
\item a docker image of the above software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}

This paper extends the work presented in \cite{mirabel-iros2016} and \cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only evoked in \cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results on several different problems.
\end{itemize}

The \paper is organized as follows. \Partie~\ref{sec:related-work} presents some related work for constrained motion planning and manipulation planning.
\Partie~\ref{partie:preliminary} introduces some preliminary notions like kinematic chains and Lie groups that are used to model the configuration space of each joint. \Partie~\ref{sec:constraints} introduces non linear constraints and solvers that are at the core of manipulation problem definition. \Partie~\ref{sec:manipulation-problem} defines the problem of prehensile manipulation in the general setting. \Partie~\ref{sec:manipulation-planning} provides a general algorithm that solves manipulation planning problems. Finally, \partie~\ref{sec:humanoid-path-planner} is devoted to the software platform implementing the notions introduced in the previous sections. Experimental results are provided for a large variety of problems.

@@

Some systems are subject to non-linear constraints. These non-linear
constraints define a sub-manifold of the configuration space the robot must
stay on. For instance, legged robots that must keep contact with the
ground and enforce quasi-static equilibrium, or multi-arm systems
grasping a same object are subject to this type of constraints. As the
volume of the constrained manifold is usually equal to zero, sampling
random configurations satisfying the constraints is an event of zero
probability. To sample configurations on the constrained manifold,
\cite{DalNakLamLau2009} and \cite{berenson2011} project random
configurations using a generalization of Newton-Raphson
algorithm. Another way to sample configurations on a constrained
manifold consists in expressing some configuration variables with
respect to others~\cite{Cortes2002}, \cite{MirLam2018} when this is
possible. \cite{atlasRRT} propose another method based on non-linear
projection. They cover the constrained manifold by growing an atlas
composed of local charts. This approximation provides a probability
distribution that is closer to the uniform distribution over the
manifold than the projection of a uniform distribution over the
configuration space. \cite{tb-rrt2016} propose a variation of the latter paper.
The main difference resides in the fact that the nodes
built on the tangent space are not immediately projected on the manifold.
\cite{CefOri2019} propose a general framework to plan task constrained motions
in the presence of moving obstacle.
\cite{KinMolKav2019} provide an in-depth review of the various approaches to
motion planning with non-linear constraints.

@@

Manipulation planning is a particular instance of path planning were
some objects are moved by some robots.
Although several instances of
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multi-contact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is constrained by
non-linear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
sub-manifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable position of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}. Some specific instances of the
problem have even been addressed recently \cite{vendittelli-2018}.

The first attempt to solve manipulation planning problems using random sampling
was proposed by \cite{simeon2004ijrr} where a reduction property reduces the
complexity of the problem.

@@

Navigation Among Movable Obstacles (NAMO) \cite{wilfong1988motion}, \cite{stilman2008planning} consists in finding
a path for a robot that needs to move objects in order to reach a goal
configuration. The final position of the objects does not matter in this case.

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying assumption is the existence of a monotone solution, that is
a solution where each object is grasped at most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic or task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{18-toussaint-RSS}.

@@

\cite{schmitt19icra} propose an approach where two robots manipulate an object in a dynamic environment. The output of the algorithm is a sequence of controllers rather than a sequence of paths.

Our work shares many ideas with \cite{HauNgt2011} where the notion of constraint
graph is present, although not as clearly expressed as in this paper. The main
contribution of our work with respect to this latter paper is that the
constraint graph is built automatically at the cost of a more restricted range
of applications. We address indeed only prehensile manipulation.

@@

Open-source software platforms are an important tool to make possible fair
comparison between algorithms.
Several software platforms are available for motion planning and/or manipulation
planning in the robotics community. The most popular one is undoubtedly OMPL \cite{ompl2012}. OMPL integrates many randomized path planning algorithms and is widely used for teaching. Recently, \cite{KinMolKav2019} proposed an extension for systems subject to non-linear constraints.

@@

The joint configuration spaces enumerated in the previous paragraph: $\mathbb{R}^n$, $SO(n)$, and $SE(n)$ are all Lie groups. The group operation is $+$ for $\mathbb{R}^n$, and composition denoted as $"."$ for $SE(n)$. We refer to~\cite{LMS94} -- Appendix A for a thorough definition of Lie groups. We expose here only properties that are useful for the following developments.

@@

Elements of $SE(2)$ are represented by a vector of dimension 4. The 2 first components represent the image of the origin. The 2 last components represent the cosine and sine of the rotation angle. Therefore the homogeneous matrix associated to $\conf=(q_1,q_2,q_3,q_4)$ is
$$
M = \left(\begin{array}{lll}
  q_3 & -q_4 & q_1 \\
  q_4 &  q_3 & q_2 \\
  0   &  0  & 1
\end{array}\right).
$$
Table~\ref{tab:vector representation} compiles this information.
\begin{table}
  \begin{tabular}{|l|c|c|}
    \hline
    Lie group type & configuration & velocity\\
    \hline
    $SE(3)$ & $(x_1,x_2,x_3,p_1,\cdots,p_4)$ & $\dconf=(\mathbf{v},\omega)$ \\
            & $\in\reals^7$                 & $\in\reals^6$ \\
    $SE(2)$ & $(x_1,x_2,\cos\theta,\sin\theta)$ & $\dconf=(\mathbf{v},\dot{\theta})$\\
            & $\in\reals^4$                 & $\in\reals^3$ \\
    $SO(3)$ & $(p_1,p_2,p_3,p_4)\in\reals^4$ & $\dconf=\omega\in\reals^3$\\
    $SO(2)$ & $(\cos\theta,\sin\theta)\in\reals^2$ & $\dconf=\dot{\theta}\in\reals$\\
    \hline
  \end{tabular}
  \caption{Main Lie group types and their vector representations. Notice that the dimensions of the configuration representation and of the velocity representation may differ.}
  \label{tab:vector representation}
\end{table}

@@

As expressed earlier, following a constant velocity\footnote{More precisely, following the vector field generated by $\dot{\conf}\in T_{\neutral}\liegroup$ according to the Lie group structure} $\dot{\conf}$ from the neutral element of a joint configuration space leads to another configuration denoted as
$$
\conf = \exp(\dconf).
$$
In some cases, we may specify in subscript the Lie group that is used: \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/explog.hpp#L34}{$\exp_{SO(3)}$}, \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/explog.hpp#L236}{$\exp_{SE(3)}$}.

For all Lie groups $\reals$, $SO(n)$, $SE(n)$, the exponential map is surjective. This means that for any $\conf\in\liegroup$, there exists $\mathbf{v}\in T_{\neutral}\liegroup$, such that $\conf = \exp(\mathbf{v})$.
Although $\exp$ is not injective, choosing the smallest norm $\mathbf{v}$ uniquely defines function $\log$ from $\liegroup$ to $T_{\neutral}\liegroup$, up to some singularities where several candidates $\mathbf{v}$ are of equal norms.
Again, we may specify the Lie group that is used: \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/log.hxx#L112}{$\log_{SE(3)}$}, \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/log.hxx#L15}{$\log_{SO(3)}$}

@@

Given a kinematic chain with joints $(J_1, \cdots, J_{njoints})$, ordered in such
a way that each joint has an index bigger than its parent in the tree, the configuration space of the robot is the Cartesian product of the joint configuration spaces.
$$
\CS \triangleq \CS_{J_1}\times \cdots \times \CS_{J_{njoints}}.
$$
$\CS$ naturally inherits the Lie group structure of the joint configuration spaces through the Cartesian product.
We denote by $nq_i$, $nv_i$ the sizes of the configuration and velocity vector representations of joint $J_i$, as defined in Table~\ref{tab:vector representation}.
The configuration and velocity of the robot can thus be represented by vectors of size $nq$ and $nv$ such that
$$
nq = \sum_{i=1}^{njoints} nq_i, \ \ \ nv = \sum_{i=1}^{njoints} nv_i
$$
We denote by $iq_i$, and $iv_i$ the indices of joint $i$ in the robot configuration and velocity vectors.
$$
iq_i = \sum_{j=1}^{i-1} nq_j\ \ \ \ iv_i = \sum_{j=1}^{i-1} nv_j
$$
With these definitions and notation, the linear interpolation between two robot
configurations $\conf_0$ and $\conf_1$ is naturally written:
$$
\conf(t) = \conf_0 \oplus t (\conf_1 - \conf_0)
$$
This formula generalizes the linear interpolation to robots with free-flying bases, getting rid of singularities of roll -- pitch -- yaw parameterization.
%% begin impl
Cartesian products of Lie groups are represented by Class \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupSpace.html}{\texttt{LiegroupSpace}}. Elements of these spaces are represented by classes
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElement}}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementRef}}, and
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementConstRef}}.
  \end{itemize}
%% end impl

@@

\begin{definition}\textbf{Non-linear constraint.} A non linear constraint is defined by a differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and is written
\begin{equation}\label{eq:constraint}
h(\conf) = 0.
\end{equation}
\end{definition}
If the robot is subject to several numerical constraints, $h_1,\cdots,h_k$ with values in $\reals^{m_1}\cdots\reals^{m_k}$, these constraints are equivalent to a single constraint $h$ with values in $\reals^{m}$, where $m=\sum_{i=1}^k m_i$, such that
$$
h(\conf) \triangleq \left(\begin{array}{c} h_1(\conf) \\ \vdots \\ h_k(\conf)
\end{array}\right).
$$
It may be useful to use a non zero right hand side for the same function $h$. We define for that parameterized non-linear constraints.
\begin{definition}\textbf{Parameterized non-linear constraint.}
  A parameterized non-linear constraint is defined by a differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and by a vector $\mathbf{h}_0$ of $\reals^m$ and is written
  $$
  h(\conf) = \mathbf{h}_0.
  $$
\end{definition}
%% begin impl
Differentiable mappings are represented by abstract Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}}.
%% end impl

@@

In this \paper, we will make use of the term Jacobian in a generalized way.
If $h$ is a differentiable function from a Lie group $\liegroup_1$ to a Lie group $\liegroup_2$, and $\conf_1$ an element of $\liegroup_1$, we will denote by $\frac{\partial h}{\partial\conf}(\conf_1)$ the operator that maps velocities in $T_{\conf_1}\liegroup_1$ to the velocity in $T_{h(\conf_1)}\liegroup_2$ \textit{transported by} $h$\footnote{If $\dconf\in T_{\conf_1}\liegroup_1$ is a velocity along a time parameterized curve $\gamma$, $\frac{\partial h}{\partial\conf}(\conf_1)\dconf$ is the velocity at
  the same time along Curve $h(\gamma)$.}.

@@

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the sub-manifold defined by the constraint} and is performed by a Gauss-Newton solver~\cite[Chapter~10]{NoceWrig06} that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}-\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i-\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{He has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ exactly solves the linear approximation, but it may not be the best choice in general.

>>>>

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{figures/example.png}
    \def\svgwidth{\columnwidth}
    \graphicspath{{./figures/}}
    \input{figures/example-graph.pdf_tex}
  \end{center}
  \caption{{\color{blue} Example of manipulation planning problem.} Top: two UR3 robots with one gripper each (X=red, Y=green, Z=blue) manipulating a cylinder with two handles. The environment contains one rectangular contact surface (in red). The cylinder has two rectangular contact surfaces (in green).
  Bottom: the corresponding constraint graph. Names of states follow Expression~(\ref{eq:state}): for instance, $(\emptyset,1)$ means that gripper of robot 2 grasps handle 1 of the cylinder. In this state, there is no placement constraint.}
  \label{fig:2ur3-cylinder}
\end{figure}

@@

This \paper therefore describes a software platform called {Humanoid Path
  Planner} taylored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modeling of prehensile manipulation based on
  non-linear constraints,
\item an original solver for non-linear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state-of-the-art development tools and methods.
\item a docker image of the above software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}
{\color{blue} Installation instructions can be found by following this \href{https://humanoid-path-planner.github.io/hpp-doc}{link}.}
This paper extends the work presented in {\color{blue} previous papers }\cite{mirabel-iros2016},\cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only evoked in {\color{blue}Mirabel \textit{et al}}~\cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results on several different problems.
\end{itemize}

The \paper is organized as follows. \Partie~\ref{sec:related-work} presents some related work for constrained motion planning and manipulation planning.
\Partie~\ref{partie:preliminary} introduces some preliminary notions like kinematic chains and Lie groups that are used to model the configuration space of each joint. \Partie~\ref{sec:constraints} introduces non-linear constraints and solvers that are at the core of manipulation problem definition. \Partie~\ref{sec:manipulation-problem} defines the problem of prehensile manipulation in the general setting. \Partie~\ref{sec:manipulation-planning} provides a general algorithm that solves manipulation planning problems. Finally, \partie~\ref{sec:humanoid-path-planner} is devoted to the software platform implementing the notions introduced in the previous sections. Experimental results are provided for a large variety of problems.

@@

Some systems are subject to non-linear constraints. These non-linear
constraints define sub-manifolds of the configuration space the robot
must stay on. For instance, legged robots that must keep contact with
the ground and enforce quasi-static equilibrium, or multi-arm systems
grasping a same object are subject to this type of constraints. As the
volume of the constrained manifold is usually equal to zero, sampling
random configurations satisfying the constraints is an event of zero
probability. To sample configurations on the constrained manifold,
{\color{blue}Dalibard \textit{et al}}~\cite{DalNakLamLau2009} and {\color{blue}Benrenson
\textit{et al}}~\cite{berenson2011} project random configurations using
a generalization of Newton-Raphson algorithm. Another way to sample
configurations on a constrained manifold consists in expressing some
configuration variables with respect to others~\cite{Cortes2002},
\cite{MirLam2018} when this is possible. {\color{blue}Jaillet \textit{et al}}~\cite{atlasRRT} propose
another method based on non-linear projection. They cover the
constrained manifold by growing an atlas composed of local
charts. This approximation provides a probability distribution that is
closer to the uniform distribution over the manifold than the
projection of a uniform distribution over the configuration
space. {\color{blue}Beobkyoon \textit{et al}}~\cite{tb-rrt2016} propose a variation of the latter paper.  The
main difference resides in the fact that the nodes built on the
tangent space are not immediately projected on the manifold.
{\color{blue}Cefalo \textit{et al}}~\cite{CefOri2019} propose a general framework to plan task constrained
motions in the presence of moving obstacle. {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019}
provide an in-depth review of the various approaches to motion
planning with non-linear constraints.

@@

Manipulation planning is a particular instance of path planning were
some objects are moved by some robots.
Although several instances of the
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multi-contact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is constrained by
non-linear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
sub-manifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable pose of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}. Some specific instances of the
problem have even been addressed recently \cite{vendittelli-2018}.

The first attempt to solve manipulation planning problems using random sampling
was proposed by {\color{blue}Siméon \textit{et al}}~\cite{simeon2004ijrr} where a reduction property reduces the
complexity of the problem.

@@

Navigation Among Movable Obstacles (NAMO) \cite{wilfong1988motion}, \cite{stilman2008planning} consists in finding
a path for a robot that needs to move objects in order to reach a goal
configuration. The final poses of the objects do not matter in this case.

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying assumption is the existence of a monotone solution, that is
a solution where each object is grasped at most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic or task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{LozKae2014}, \cite{18-toussaint-RSS}.

@@

{\color{blue}Schmitt \textit{et al}}~\cite{schmitt19icra} propose an approach where two robots manipulate an object in a dynamic environment. The output of the algorithm is a sequence of controllers rather than a sequence of paths.

Our work shares many ideas with {\color{blue}Hauser and Ng-Thow-Hing} \cite{HauNgt2011} where the notion of constraint
graph is present, although not as clearly expressed as in this paper. The main
contribution of our work with respect to this latter paper is that the
constraint graph is built automatically at the cost of a more restricted range
of applications. We address indeed only prehensile manipulation.

@@

Open-source software platforms are an important tool to make possible fair
comparison between algorithms.
Several software platforms are available for motion planning and/or manipulation
planning in the robotics community. The most popular one is undoubtedly OMPL \cite{ompl2012}. OMPL integrates many randomized path planning algorithms and is widely used for teaching. Recently, {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019} proposed an extension for systems subject to non-linear constraints.

@@

The joint configuration spaces enumerated in the previous paragraph: $\mathbb{R}^n$, $SO(n)$, and $SE(n)$ are all Lie groups. The group operation is $+$ for $\mathbb{R}^n$, and composition denoted as $"."$ for $SE(n)$. We refer to~{\color{blue}Murray \textit{et al}}~\cite[Appendix A]{LMS94} for a thorough definition of Lie groups. We expose here only properties that are useful for the following developments.

@@

Elements of $SE(2)$ are represented by a vector of dimension 4. The 2 first components represent the image of the origin. The 2 last components represent the cosine and sine of the rotation angle. Therefore the homogeneous matrix associated to $\conf=(q_1,q_2,q_3,q_4)$ is
$$
M = \left(\begin{array}{lll}
  q_3 & -q_4 & q_1 \\
  q_4 &  q_3 & q_2 \\
  0   &  0  & 1
\end{array}\right).
$$
Table~\ref{tab:vector representation} compiles this information.
\begin{table}
  \begin{tabular}{|l|c|c|}
    \hline
    Lie group type & configuration & velocity\\
    \hline
    $SE(3)$ & $(x_1,x_2,x_3,p_1,\cdots,p_4)$ & $\dconf=(\mathbf{v},\omega)$ \\
            & $\in\reals^7$                 & $\in\reals^6$ \\
    $SE(2)$ & $(x_1,x_2,\cos\theta,\sin\theta)$ & $\dconf=(\mathbf{v},\dot{\theta})$\\
            & $\in\reals^4$                 & $\in\reals^3$ \\
    $SO(3)$ & $(p_1,p_2,p_3,p_4)\in\reals^4$ & $\dconf=\omega\in\reals^3$\\
    $SO(2)$ & $(\cos\theta,\sin\theta)\in\reals^2$ & $\dconf=\dot{\theta}\in\reals$\\
    \hline
  \end{tabular}
  \caption{Main Lie group types and their vector representations. Notice that the dimensions of the configuration representation and of the velocity representation may differ. {\color{blue} Using $(\cos\theta,\sin\theta)$ instead of $\theta$ for $SO(2)$ and $SE(2)$ makes the parameterization continuous when $\theta$ discontinuously switches from $-\pi$ to $\pi$}}
  \label{tab:vector representation}
\end{table}

@@

As expressed earlier, following a constant velocity\footnote{More precisely, following the vector field generated by $\dot{\conf}\in T_{\neutral}\liegroup$ according to the Lie group structure} $\dot{\conf}$ from the neutral element of a joint configuration space leads to another configuration denoted as
$$
\conf = \exp(\dconf).
$$
In some cases, we may specify in subscript the Lie group that is used: \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/explog.hpp\#L34}{$\exp_{SO(3)}$}, \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/explog.hpp\#L236}{$\exp_{SE(3)}$}.

For all Lie groups $\reals$, $SO(n)$, $SE(n)$, the exponential map is surjective. This means that for any $\conf\in\liegroup$, there exists $\mathbf{v}\in T_{\neutral}\liegroup$, such that $\conf = \exp(\mathbf{v})$.
Although $\exp$ is not injective, choosing the smallest norm $\mathbf{v}$ uniquely defines function $\log$ from $\liegroup$ to $T_{\neutral}\liegroup$, up to some singularities where several candidates $\mathbf{v}$ are of equal norms.
Again, we may specify the Lie group that is used: \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/log.hxx\#L112}{$\log_{SE(3)}$}, \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/log.hxx\#L15}{$\log_{SO(3)}$}

@@

Given a kinematic chain with joints $(J_1, \cdots, J_{njoints})$, ordered in such
a way that each joint has an index bigger than its parent in the tree, the configuration space of the robot is the Cartesian product of the joint configuration spaces.
$$
\CS \triangleq \CS_{J_1}\times \cdots \times \CS_{J_{njoints}}.
$$
$\CS$ naturally inherits the Lie group structure of the joint configuration spaces through the Cartesian product.
We denote by $nq_i$, $nv_i$ the sizes of the configuration and velocity vector representations of joint $J_i$, as defined in Table~\ref{tab:vector representation}.
The configuration and velocity of the robot can thus be represented by vectors of size $nq$ and $nv$ such that
$$
nq = \sum_{i=1}^{njoints} nq_i, \ \ \ nv = \sum_{i=1}^{njoints} nv_i
$$
We denote by $iq_i$, and $iv_i$ the {\color{blue}starting} indices of joint $i$ in the robot configuration and velocity vectors.
$$
iq_i = \sum_{j=1}^{i-1} nq_j\ \ \ \ iv_i = \sum_{j=1}^{i-1} nv_j
$$
With these definitions and notation, the linear interpolation between two robot
configurations $\conf_0$ and $\conf_1$ is naturally written:
$$
\conf(t) = \conf_0 \oplus t (\conf_1 {\color{blue}\ominus} \conf_0)
$$
This formula generalizes the linear interpolation to robots with free-flying bases, getting rid of singularities of roll -- pitch -- yaw parameterization.
%% begin impl
Cartesian products of Lie groups are represented by Class \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupSpace.html}{\texttt{LiegroupSpace}}. Elements of these spaces are represented by classes
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElement}}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementRef}}, and
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementConstRef}}.
  \end{itemize}
%% end impl

@@

\begin{definition}\textbf{Non-linear constraint.} A non-linear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and is written
\begin{equation}\label{eq:constraint}
h(\conf) = 0.
\end{equation}
\end{definition}
If the robot is subject to several numerical constraints, $h_1,\cdots,h_k$ with values in $\reals^{m_1}\cdots\reals^{m_k}$, these constraints are equivalent to a single constraint $h$ with values in $\reals^{m}$, where $m=\sum_{i=1}^k m_i$, such that
$$
h(\conf) \triangleq \left(\begin{array}{c} h_1(\conf) \\ \vdots \\ h_k(\conf)
\end{array}\right).
$$
It may be useful to use a non zero right hand side for the same function $h$. We define for that parameterized non-linear constraints.
\begin{definition}\textbf{Parameterized non-linear constraint.}
  A parameterized non-linear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and by a vector $\mathbf{h}_0$ of $\reals^m$ and is written
  $$
  h(\conf) = \mathbf{h}_0.
  $$
\end{definition}
%% begin impl
{\color{blue} Piece-wise} differentiable mappings are represented by abstract Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}}.
%% end impl

@@

In this \paper, we will make use of the term Jacobian in a generalized way.
If $h$ is a {\color{blue} piece-wise} differentiable function from a Lie group $\liegroup_1$ to a Lie group $\liegroup_2$, and $\conf_1$ an element of $\liegroup_1$, we will denote by $\frac{\partial h}{\partial\conf}(\conf_1)$ the operator that maps velocities in $T_{\conf_1}\liegroup_1$ to the velocity in $T_{h(\conf_1)}\liegroup_2$ \textit{transported by} $h$\footnote{If $\dconf\in T_{\conf_1}\liegroup_1$ is a velocity along a time parameterized curve $\gamma$, $\frac{\partial h}{\partial\conf}(\conf_1)\dconf$ is the velocity at
  the same time along Curve $h(\gamma)$.}.

@@

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the sub-manifold defined by the constraint} and is performed by a Gauss-Newton solver~\cite[Chapter~10]{NoceWrig06} that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}{\color{blue}\ominus}\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i{\color{blue}\ominus}\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{He has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ exactly solves the linear approximation, but it may not be the best choice in general.

The computation of $\alpha_i$ is performed by a line search algorithm.
The algorithm stops when the norm of each $h_i(\conf_{i+1})$ is below a given error threshold.
%% begin impl
Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1HierarchicalIterative.html}{\texttt{HierarchicalIterative}} implements the above Newton method.
{\color{blue}
Several line search methods are implemented:
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1Backtracking.html}{\texttt{Backtracking}} \cite{backtracking},
\item \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1ErrorNormBased.html}{\texttt{ErrorNormBased}}:
  $$\alpha_{i} = C - K  \text{tanh}(a \frac{\|f(\mathbf{q}_i)\|}{\epsilon^2} + b),$$
  where $C$, $K$, $a$, and $b$ are constant values, and $\epsilon$ is the
  error threshold,
\item \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1FixedSequence.html}{\texttt{FixedSequence}} implements a fixed sequence of $\alpha_i$ that converges to 1,
\item and \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1Constant.html}{\texttt{Constant}} sets $\alpha_i$ to 1.
\end{itemize}
}
Note that to define a new constraint, the user needs to derive class \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}} and to implement methods \texttt{impl\_compute} and \texttt{impl\_jacobian}.
%% end impl
